---
title: "Optimization Solvers in PortfolioAnalytics"
vignette: >
  %\VignetteIndexEntry{Optimization Solvers in PortfolioAnalytics}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-overflow: wrap
    df-print: kable
  pdf:
    toc: true
    toc-depth: 3
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| include: false
library(PortfolioAnalytics)
library(PerformanceAnalytics)
```

## Introduction

PortfolioAnalytics provides a flexible framework for portfolio optimization
that separates the *specification* of objectives and constraints from the
*solver* used to find optimal portfolios. This solver-agnostic design allows
users to swap optimization backends without rewriting their portfolio
specification, making it straightforward to compare results across methods.

This vignette provides:

- An overview of all supported optimization solvers
- A compatibility matrix mapping objective specifications to solvers
- Worked examples for each objective/solver combination
- Performance benchmarks comparing time-to-converge and solution quality
- Guidance on solver selection

## Solver Overview

PortfolioAnalytics supports solvers spanning three categories: convex/closed-form
solvers that exploit problem structure for exact solutions, metaheuristic solvers
that use stochastic search for arbitrary objectives, and multi-objective solvers
for Pareto optimization.

### Convex / Closed-Form Solvers

#### CVXR

CVXR provides the most comprehensive closed-form solver support in
PortfolioAnalytics. It uses disciplined convex programming (DCP) to
model and solve LP, QP, and SOCP problems, supporting variance, ES/CVaR,
CSM, EQS, and HHI objectives natively.

<!-- TODO: CVXR sub-solvers (OSQP, CLARABEL, SCS, ECOS, etc.) -->

#### ROI

The R Optimization Infrastructure (ROI) dispatches to plugin solvers:
`quadprog` for QP problems (min variance, max quadratic utility) and
`glpk`/`symphony` for LP/MILP problems (max return, min ES/CVaR).

#### osqp

A standalone QP solver supporting mean and variance/StdDev objectives.
Fast for pure quadratic problems but limited in objective scope.

#### Rglpk

A standalone LP/MILP solver supporting mean and ES/CVaR objectives.

### Metaheuristic Solvers

These solvers accept *any* R function as an objective, making them suitable
for non-convex or black-box problems. The trade-off is longer run times and
no guarantee of global optimality.

- **DEoptim** -- Differential Evolution
- **GenSA** -- Generalized Simulated Annealing
- **pso** -- Particle Swarm Optimization
- **random** -- Random portfolio search (built-in)

### Multi-Objective

- **mco** -- NSGA-II for Pareto-optimal frontiers across multiple objectives

<!-- TODO: solver summary table -->

## Objective & Constraint Support Matrix

<!-- TODO: master compatibility matrix -->
<!-- Rows: maxret, minvar, minES, minCSM, maxSR, maxSTARR, CSMratio, -->
<!--        EQSratio, maxQU, riskbudget, weight_concentration            -->
<!-- Columns: CVXR, ROI, osqp, Rglpk, DEoptim, GenSA, pso, random, mco  -->

## Worked Examples

### Data Setup

```{r}
#| label: data-setup
#| eval: false
data(edhec)
R <- edhec["2008::2012", 1:6]

funds <- colnames(R)
port <- portfolio.spec(assets = funds)
port <- add.constraint(port, type = "full_investment")
port <- add.constraint(port, type = "long_only")
```

### Minimum Variance

<!-- TODO: CVXR, ROI, osqp, DEoptim, GenSA, pso, random -->

### Minimum ES / CVaR

<!-- TODO: CVXR, ROI, Rglpk, DEoptim, GenSA, pso, random -->

### Minimum CSM

<!-- TODO: CVXR (native), DEoptim, GenSA, pso, random (manual) -->

### Maximum Return

<!-- TODO: CVXR, ROI, Rglpk, DEoptim, GenSA, pso, random -->

### Maximum Sharpe Ratio

<!-- TODO: CVXR, ROI, DEoptim, GenSA, pso, random -->

### Maximum STARR

<!-- TODO: CVXR, ROI, DEoptim, GenSA, pso, random -->

### Maximum Quadratic Utility

<!-- TODO: ROI, CVXR, DEoptim, GenSA, pso, random -->

### Risk Budgets

<!-- TODO: DEoptim, GenSA, pso, random -->

### Weight Concentration (HHI)

<!-- TODO: CVXR, ROI, DEoptim, GenSA, pso, random -->

## Performance Benchmarks

<!-- TODO: methodology, timing tables, convergence quality, visuals -->

## Solver Selection Guide

<!-- TODO: decision tree, scalability notes -->

## Gaps & Normalization Notes

<!-- TODO: identified gaps, inconsistencies, planned improvements -->

## Appendix

### Constraint Type Reference

<!-- TODO: full constraint reference table -->

### Session Info

```{r}
#| label: session-info
#| eval: false
sessionInfo()
```
